---
title: "PSY4960- Advanced Topics "
author: "Amanda Mae Woodward"
date: "2022-11-28"
output: html_document
---
```{r loadLibraries, include= FALSE}
#echo - hides code, gives output
#include- hides code, hides output
library(tidyverse)
library(readr)
```

```{r importData}
worldCupDat <- read_csv("wc_matches.csv")
#View(worldCupDat)
```

```{r checkClass}
summary(worldCupDat)
worldCupDat$team1<- as.factor(worldCupDat$team1)
worldCupDat$team2<- as.factor(worldCupDat$team2)
```

### Data set Manipulation
-	Data manipulation (wide and long format) 
```{r datamanipulationPrep}
worldCupDat$gameNumber<- 1:48
```

```{r pivotLonger}
names(worldCupDat)
worldCupLong<- pivot_longer(worldCupDat, cols= c("team1", "team2"), values_to = "nameOfTeam", names_to= "homeAway")
```

- Missing values
```{r}
worldCupBad<- na.omit(worldCupDat)
summary(is.na(worldCupDat))
#na.rm=TRUE in the analysis
```
-	Separate function
```{r sepReview}
"data %>% separate(column to separate, what you want to call new columns, sep = thing that separates it)"
names(worldCupDat)
separatedDate<- worldCupDat %>% separate(date, into= c("month", "day", "year"), sep = "/")
```

-	What’s a subset and how to combine them 
```{r}
byTeam<-split(worldCupDat, worldCupDat$team1)
#View(byTeam$Argentina)

Ateam<- rbind.data.frame(byTeam$Argentina, byTeam$Australia)
```

-	Merging data and getting rid of duplicates
```{r}
Argentina<- rbind.data.frame(byTeam$Argentina, byTeam$Argentina)
noDuplicates<-unique(Argentina)
```
###Tables
-	Tables in R
```{r}
table(worldCupDat$score1)
library(expss)
cro(worldCupDat$score1)
cro(worldCupDat$team1, worldCupDat$score1)
```

### Statistics
-	Omitting missing data 
```{r}
mean(worldCupDat$score1)
mean(worldCupDat$score1, na.rm=TRUE)
```
-	Repeated measures anova
```{r}
names(worldCupLong)
summary(aov(score1~ gameNumber+ nameOfTeam, data= worldCupLong))
```

-	When to use nonparametric analyses
-	Fisher’s exact tests
-	Interactions and ANOVA
-	Logistic regression
```{r}
glm(DV ~ IV, data= dataset, family="binomial")
```


### Graphing
-	More ggplot() (interactions, position dodge())
-	Visualizations for different tests

### Coding Syntax
-	Pipes 

### specific codes
-	For loops/ if else statements
-	Knitting










